PLUNGER: Reproducing FLUSH+RELOAD: A High-Resolution, Low-Latency Side Channel Attack On GnuPG

### 2014-12-17 paper/paper.tex

A typical cache architecture is laid out in three levels, generally referred to
as the \textit{L1}, \textit{L2}, and \textit{L3}. The L3 cache is usually larger
than than the L2, which itself is usually larger than the L1; L1 and L2 caches
are also generally allocated on a per-core basis in the case of multi-core CPUs,
whereas the L3 cache is usually shared between all cores.  L1 caches are also
typically divided into \textit{instruction} sections and \textit{data} sections,
each reserved for that particular type of instruction.  It is equally important
for instruction memory as well as data memory to be cached, as both require fast
access in order for the CPU to be performant. For example, the Intel Haswell i7
4600U CPU has a 128KiB L1 cache, divided into two 64KiB caches for instructions
and data; a 512KiB L2 cache, and a 4MiB L3 cache.

@@

Windows, Linux, and possibly OS X\footnote{It is unclear whether or not OS X
implements page deduplication, but as far as we can tell it does} also perform a
more aggressive form of page sharing known as \textit{page deduplication}. With
this scheme, the operating system scans memory pages during idle CPU cycles, and
looks for pages that are identical in content. It collapses those pages into a
single page that is then shared by all processes that were accessing those
identical pages. As a result of this page deduplication scheme,
processes may be unknowingly or unintentionally sharing memory with one another.
However, to prevent malicious access to the shared pages, the operating system
maps shared pages as copy-on-write; that is, when a process wishes to write to a
page, it must first make a local copy for itself that it can safely modify. When
a write is requested, the CPU traps to the kernel, which then remaps a copy of
the desired page for the process requesting the write, and then returns control
to the calling process. Read operations in this scheme are safe, as they do not
modify the targeted page.

@@

The \flushreload{} attack takes advantage of page deduplication on modern
processors. The basic attack works as follows. The attacker loads into memory
some block of data that it believes the victim to also be accessing. The system
then deduplicates these two identical pages. The attacker can then use the
instruction \texttt{clflush} to flush memory lines in that block from the CPU caches. The
attacker then waits for some time $T$ and attempts to access a memory line
in this page, and measures the time (in cycles) it takes for the line to be
loaded. If the memory line was accessed by the victim during $T$, the time taken
will be shorter because it will be cached, and if it was not accessed, the time taken will be longer.
Consequently, the attacker can determine if the line was accessed during $T$.

@@

Arch Linux provided the poorest results by far. No meaningful data were
discernible under any configuration because there was too much timing noise
during the spying, which, in the case of the Figure~\ref{fig:data-arch}, occurred roughly
between timeslots 7000 and 18000 on the x-axis. Given that there were no
significant differences between this and the configuration running on OS X, we
were unsure why this kept occurring. The most obvious
difference between the two systems is that GnuPG on Arch Linux was compiled
under GCC 4.9.2 instead of GCC 4.2, as it was on OS X; however, it was not clear how
this would change the output so drastically, or if it did at all. Our two
hypotheses for this behaviour were that Arch Linux
was being run as a guest OS on a Windows 8.1 host, and that it was being run on
a Haswell-series Intel i7, which could have affected GCC's compilation of GnuPG
because this CPU series
did not exist at the publication time of the original paper.  However, both of
these hypotheses were disproven by testing on a second machine running Arch
Linux as a host OS on an older Intel Celeron CPU, in which the attack produced
the same noisy results.

@@

Many of our preliminary results were plagued by a lot of noise, much of which
resulted from speculative execution. To reduce the effects of this, we attempted
to probe the very close to the ends of the respective functions. We found that
simply setting breakpoints to the relevant lines in GDB would not yield good
addresses to probe. Even setting a breakpoint to the last line of the function
in the source would not return an address of an instruction close to the return.
To find the best addresses, we disassembled the GPG binary using
\texttt{objdump} and located addresses near each function's \texttt{retq}
instruction.

>>>>

A typical cache architecture is laid out in three levels, generally referred to
as the \textit{L1}, \textit{L2}, and \textit{L3}. The L3 cache is usually larger
than than the L2, which itself is usually larger than the L1; L1 and L2 caches
are also generally allocated on a per-core basis in the case of multi-core CPUs,
whereas the L3 cache is usually shared between all cores.  L1 caches are also
typically divided into \textit{instruction} sections and \textit{data} sections,
each reserved for that particular type of instruction.  It is equally important
for instruction memory as well as data memory to be cached, as both require fast
access in order for the CPU to be performant. For example, the Intel Haswell i7
4600U CPU has a 128KiB L1 cache, divided into two 64KiB caches for instructions
and data; a 512KiB L2 cache; and a 4MiB L3 cache.

@@

Windows, Linux, and possibly OS X\footnote{It is unclear whether or not OS X
implements page deduplication, but as far as we can tell it does} also perform a
more aggressive form of page sharing known as \textit{page deduplication}. With
this scheme, the operating system scans memory pages during idle CPU cycles, and
looks for pages that are identical in content. It collapses those pages into a
single page that is then shared by all processes that were accessing those
identical pages. As a result of this page deduplication scheme,
processes may be unknowingly or unintentionally sharing memory with one another.
However, to prevent malicious access to the shared pages, the operating system
maps shared pages as copy-on-write; that is, when a process wishes to write to a
page, it must first make a local copy for itself that it can safely modify. When
a write is requested, the CPU traps to the kernel, which then remaps a copy of
the desired page for the process requesting the write, and then returns control
to the calling process. Read operations in this scheme do not require such a trap
as they do not modify the targeted page and are therefore ``safe''.

@@

The \flushreload{} attack takes advantage of page deduplication on modern
processors. The basic attack works as follows: the attacker loads into memory
some block of data that it believes the victim to also be accessing. The system
then deduplicates these two identical pages. The attacker can then use the
instruction \texttt{clflush} to flush memory lines in that block from the CPU caches. The
attacker then waits for some time $T$ and attempts to access a memory line
in this page, and measures the time (in cycles) it takes for the line to be
loaded. If the memory line was accessed by the victim during $T$, the time taken
will be shorter because it will be cached, and if it was not accessed, the time taken will be longer.
Consequently, the attacker can determine if the line was accessed during $T$.

@@

Arch Linux provided the poorest results by far. No meaningful data were
discernible under any configuration because there was too much timing noise
during the spying, which, in the case of the Figure~\ref{fig:data-arch}, occurred roughly
between timeslots 7000 and 18000 on the x-axis. Given that there were no
significant differences between this and the configuration running on OS X, we
were unsure why this kept occurring. The most obvious
difference between the two systems is that GnuPG on Arch Linux was compiled
under GCC 4.9.2 instead of GCC 4.2, as it was on OS X; however, it was not clear how
this would change the output so drastically, or if it did at all.

Our two hypotheses for this different behaviour were that Arch Linux
was being run as a guest OS on a Windows 8.1 host, and that it was being run on
a Haswell-series Intel i7, which could have affected GCC's compilation of GnuPG
because this CPU series
did not exist at the publication time of the original paper.  However, both of
these hypotheses were disproven by testing on a second machine running Arch
Linux as a host OS on an older Intel Celeron CPU, in which the attack produced
the same noisy results.

Additionally, it may be the case that the addresses determined with GDB were inaccurate,
and that probing based on addresses determined via \texttt{objdump} would produce
better results.

@@

Many of our preliminary results were plagued by a lot of noise, much of which
resulted from speculative execution. To reduce the effects of this, we attempted
to probe the very close to the ends of the respective functions. We found that
simply setting breakpoints to the relevant lines in GDB would not yield good
addresses to probe. Even setting a breakpoint on the last line of the function
in the source would not return an address of an instruction close to the return.
To find the best addresses, we disassembled the GPG binary using
\texttt{objdump} and located addresses near each function's \texttt{retq}
instruction.
